\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-07-01 doc/src/IMSIC.tex
Upgrade chapter status; remove Duo-PLIC chapter to a separate document
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

\textbf{%
This chapter is considered stable but could still change before being
ratified as standard by the {\RISCV} International Association.%
}
\bigskip

@@

This section describes the registers of an interrupt file that
are accessed indirectly through a \z{*iselect} CSR (\z{miselect},
\z{siselect}, or \z{vsiselect}) and its partner \z{*ireg} CSR
(\z{mireg}, \z{sireg}, or \z{vsireg}).

For each interrupt file, the width of all indirectly accessed registers
is the width of the \z{*ireg} CSR used to access those registers,
either MXLEN, SXLEN, or VSXLEN.

@@

When the interrupt file's registers are 32~bits (as determined by
the XLEN for the corresponding privilege level), register \z{eip}$k$
contains the pending bits for interrupts with identity numbers
$k\times\mbox{32}$ through ${k\times\mbox{32} + \mbox{31}}$.
For an implemented interrupt identity $i$ within that range, the
pending bit for interrupt~$i$ is bit $(i\bmod\mbox{32})$ of \z{eip}$k$.

When the interrupt file's registers are 64~bits, the odd-numbered
registers \z{eip1}, \z{eip3}, \dots \z{eip63} do not exist.
In that case, if the \z{*iselect} CSR is an odd value in the range
\z{0x81}--\z{0xBF}, an attempt to access the matching \z{*ireg} CSR
raises an illegal instruction exception, unless done in VS-mode, in
which case it raises a virtual instruction exception.
For even~$k$, register \z{eip}$k$ contains the pending bits
for interrupts with identity numbers $k\times\mbox{32}$ through
${k\times\mbox{32} + \mbox{63}}$.
For an implemented interrupt identity~$i$ within that range, the
pending bit for interrupt~$i$ is bit $(i\bmod\mbox{64})$ of \z{eip}$k$.

@@

When the interrupt file's registers are 32~bits (as determined by
the XLEN for the corresponding privilege level), register \z{eie}$k$
contains the enable bits for interrupts with identity numbers
$k\times\mbox{32}$ through ${k\times\mbox{32} + \mbox{31}}$.
For an implemented interrupt identity $i$ within that range, the enable
bit for interrupt~$i$ is bit $(i\bmod\mbox{32})$ of \z{eie}$k$.

When the interrupt file's registers are 64~bits, the odd-numbered
registers \z{eie1}, \z{eie3}, \dots \z{eie63} do not exist.
In that case, if the \z{*iselect} CSR is an odd value in the range
\z{0xC1}--\z{0xFF}, an attempt to access the matching \z{*ireg} CSR
raises an illegal instruction exception, unless done in VS-mode, in
which case it raises a virtual instruction exception.
For even~$k$, register \z{eie}$k$ contains the enable bits for
interrupts with identity numbers $k\times\mbox{32}$ through
${k\times\mbox{32} + \mbox{63}}$.
For an implemented interrupt identity~$i$ within that range, the enable
bit for interrupt~$i$ is bit $(i\bmod\mbox{64})$ of \z{eie}$k$.

>>>>

@@

This section describes the registers of an interrupt file that
are accessed indirectly through a \z{*iselect} CSR (\z{miselect},
\z{siselect}, or \z{vsiselect}) and its partner \z{*ireg} CSR
(\z{mireg}, \z{sireg}, or \z{vsireg}).
The width of these indirect accesses is always the current XLEN,
32~bits for RV32 code, or 64~bits for RV64 code.

@@

When the current XLEN = 32, register \z{eip}$k$
contains the pending bits for interrupts with identity numbers
$k\times\mbox{32}$ through ${k\times\mbox{32} + \mbox{31}}$.
For an implemented interrupt identity $i$ within that range, the
pending bit for interrupt~$i$ is bit $(i\bmod\mbox{32})$ of \z{eip}$k$.

When the current XLEN = 64, the odd-numbered
registers \z{eip1}, \z{eip3}, \dots \z{eip63} do not exist.
In that case, if the \z{*iselect} CSR is an odd value in the range
\z{0x81}--\z{0xBF}, an attempt to access the matching \z{*ireg} CSR
raises an illegal instruction exception, unless done in VS-mode, in
which case it raises a virtual instruction exception.
For even~$k$, register \z{eip}$k$ contains the pending bits
for interrupts with identity numbers $k\times\mbox{32}$ through
${k\times\mbox{32} + \mbox{63}}$.
For an implemented interrupt identity~$i$ within that range, the
pending bit for interrupt~$i$ is bit $(i\bmod\mbox{64})$ of \z{eip}$k$.

@@

When the current XLEN = 32, register \z{eie}$k$
contains the enable bits for interrupts with identity numbers
$k\times\mbox{32}$ through ${k\times\mbox{32} + \mbox{31}}$.
For an implemented interrupt identity $i$ within that range, the enable
bit for interrupt~$i$ is bit $(i\bmod\mbox{32})$ of \z{eie}$k$.

When the current XLEN = 64, the odd-numbered
registers \z{eie1}, \z{eie3}, \dots \z{eie63} do not exist.
In that case, if the \z{*iselect} CSR is an odd value in the range
\z{0xC1}--\z{0xFF}, an attempt to access the matching \z{*ireg} CSR
raises an illegal instruction exception, unless done in VS-mode, in
which case it raises a virtual instruction exception.
For even~$k$, register \z{eie}$k$ contains the enable bits for
interrupts with identity numbers $k\times\mbox{32}$ through
${k\times\mbox{32} + \mbox{63}}$.
For an implemented interrupt identity~$i$ within that range, the enable
bit for interrupt~$i$ is bit $(i\bmod\mbox{64})$ of \z{eie}$k$.

### 2023-07-01 doc/src/preface.tex
Improve formatting
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>
Update version number to 1.0-RC6
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

This document describes an Advanced Interrupt Architecture
for {\RISCV} systems.

No part of this document has yet been ratified
by the {\RISCV} International Association.
The table below shows the current status of each chapter,
and also indicates which chapters specify extensions to the
{\RISCV} ISA (instruction set architecture) and which are non-ISA.

{
\begin{table}[hbt]
\centering
\begin{tabular}{|l|c|c|}
\hline
Chapter                                                  & ISA? & Status \\
\hline
\hline
1.\ Introduction                                         & ---  & Frozen \\
2.\ Control and Status Registers (CSRs) Added to Harts   & Yes  & Frozen \\
3.\ Incoming MSI Controller (IMSIC)                      & Yes  & Frozen \\
4.\ Advanced Platform-Level Interrupt Controller (APLIC) & No   & Frozen \\
5.\ Interrupts for Machine and Supervisor Levels         & Yes  & Frozen \\
6.\ Interrupts for Virtual Machines (VS Level)           & Yes  & Frozen \\
7.\ Interprocessor Interrupts (IPIs)                     & No   & Frozen \\
8.\ IOMMU Support for MSIs to Virtual Machines           & No   & Frozen \\
\hline
\end{tabular}
\end{table}
}

An implementation adhering to the current document might not conform
to an eventual ratified Advanced Interrupt Architecture for {\RISCV}.
However, all chapters are \emph{frozen}, meaning they are not expected
to change significantly before being put up for ratification.

@@

@@

\section*{Changes for RC4 (Ratification Candidate 4)}

>>>>

This document describes an Advanced Interrupt Architecture
for {\RISCV} systems.
This specification was ratified by the
{\RISCV} International Association in June of 2023.

The table below indicates which chapters
of this document specify extensions to the
{\RISCV} ISA (instruction set architecture) and which are non-ISA.

{
\begin{table}[hbt]
\centering
\begin{tabular}{|l|c|}
\hline
Chapter                                                  & ISA? \\
\hline
\hline
1.\ Introduction                                         & ---  \\
2.\ Control and Status Registers (CSRs) Added to Harts   & Yes  \\
3.\ Incoming MSI Controller (IMSIC)                      & Yes  \\
4.\ Advanced Platform-Level Interrupt Controller (APLIC) & No   \\
5.\ Interrupts for Machine and Supervisor Levels         & Yes  \\
6.\ Interrupts for Virtual Machines (VS Level)           & Yes  \\
7.\ Interprocessor Interrupts (IPIs)                     & No   \\
8.\ IOMMU Support for MSIs to Virtual Machines           & No   \\
\hline
\end{tabular}
\end{table}
}

\section*{Changes for the ratified version 1.0}

Resolved some inconsistencies in Chapter~\ref{ch:CSRs}
about when to raise a virtual instruction exception
versus an illegal instruction exception.

@@

Added clarification about the term \emph{\mbox{IOMMU}}
used in Chapter~\ref{ch:IOMMU}.

@@

\section*{Changes for RC4}
